Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-05-03 LaTeX/response.tex

\section{Reviewer 1}

The turnover within a species from one genetic means of sex determination to another is surprisingly common, and it poses a challenging puzzle for evolutionary geneticists.  Existing theory on the topic suggests that there must be either sexually-antagonistic selection or selection variants that bring the sex ratio closer to 1:1.  This paper builds explicit models of the situation and asks about the role of selection in the haploid stage (i.e. meiotic drive and gametic selection), and finds that such effects can also serve as the primary driver of turnover of sex determining mechanisms.  While this is a rather focused topic, it is a striking evolutionary puzzle, and the paper presents a very satisfying solution that is much more general than we had previously thought.  It also overturns the primacy of Fisherian sex ratio evolution as the driver of the dynamics of genes of this sort.

The authors devise a satisfyingly simple genetic scenario, with one ancestral sex-determining locus, one locus under selection, and a third locus for the nascent sex determining locus.  They manage to collapse some messy algebra into terms like the ``haplotypic growth rate" that allow the reader to intuit many of the results.  I only have a few suggestions:

\textcolor{blue}{
We thank the reviewer for their careful consideration of our manuscript and pithy summary. Indeed, we have incorporated the sentiment of ``a striking evolutionary puzzle'' to help frame our study, L51. We believe that the modifications outlined below have helped to further improve the communication of our results. 
}

@@

\noindent\subsection{}
The selected locus allows for three modes of selection: gametic selection (weighting the frequencies of gamete genotypes), the usual diploid viability selection, and meiotic drive (where heterozygotes may produce gametes that deviate from 50:50 proportions).  It should be noted that the model does not accommodate fertility or sexual selection.

\noindent\paragraph{Response}
\textcolor{blue}{
One could interpret our diploid selection terms as a combination of viability and fertility selection, where fertility selection is independent of the mating partner.
However, fertility selection that depends on the mating pair would require a different formulation.
We now note this in the model section (L120-122):
``Diploid males and females then experience viability and/or individual-based fertility selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$.
We do not consider fertility selection that depends on the mating partner, e.g., sexual selection with variation in choosiness."
}

@@

\noindent\subsection{}
The ``haplotypic growth rate" plays a really essential role in providing intuition about the results. Similarly the $\chi_{mi}^{(XY)}$ terms also are crucial.  I suggest giving a short name for the chi terms (currently they are the ``rate at which mutant haplotypes on background i recombine onto the other A locus background in heterozygotes"...).   The importance of these becomes clear in Table 2, but I wonder if there is a graphical way to make their meaning even clearer.

\noindent\paragraph{Response}
\textcolor{blue}{
We now refer to the $\chi_{mi}^{(XY)}$ terms as the ``dissociative force" as they break down linkage disequilibrium (L161). 
We have attempted various graphical schemes for representing the three loci in two sexes and the associations that evolve between them in various scenarios. 
However, apart from Fig 1, we found these too complicated or opaque to be useful. 
Instead, we have substantially restructured the results section and reduced the technicality (see responses XXXXXX), which should generally help to make the important messages clearer. 
Specific to this point, in the paragraph 179-199, we itemize three take-home messages from table 2 about the ``haplotypic growth rates'' and ``dissociative forces''. 
}

@@

\noindent\subsection{}
It needs to be stated somewhere that the models are entirely deterministic, implying that the authors assume that the selective forces always exceed the force of random drift. It might be useful to discuss this a bit. Is there evidence that in no case has drift been a key factor in changes of sex determination system?  The discussion mentions alleles of weaker effect, but does not mention that in this scenario, models of weak effects would need to consider drift as well.

\noindent\paragraph{Response}
\textcolor{blue}{
This is true, we now mention that the model is deterministic, and hence ignores drift (L130-131). 
As suggested, when discussing the model limitations at the end of the Discussion we now mention a study on the effect of drift during trans-GSD transitions, which tend to drift in the direction of epistatically-dominant sex-determining loci (L538-541): ``It should be noted, however, that the dynamics of sex-determining alleles with very weak effect will be influenced by genetic drift, which itself has been shown to cause trans-GSD transitions (Veller et al 2017)."}

@@

\noindent\subsection{}
A glitch appears in the section ``Loose linkage with the ancestral sex-determining region" on line 270, where the text reads, ``To leading order in selection, the leading eigenvalues are..."

\noindent\paragraph{Response}
\textcolor{blue}{Thank you. We have dropped the words ``the leading eigenvalues are".}

@@

\noindent\subsection{}
Gametic selection among sperm is a bit different from sperm competition, since the latter involves pair-wise (or multi-way) bouts, whereas gametic selection assigns constant fitnesses to each haplotype. I would replace occurrences of ``sperm competition" with ``male gametic selection", or maybe ``sperm gametic competition". To be consistent, replace ``egg competition" with ``female/egg gametic competition".

\noindent\paragraph{Response}
\textcolor{blue}{We have made the suggested replacements.
In addition, we have added a more detailed discussion of how gametic competition relates to the number of mating partners involved, L108-118 (see also response XXXXXX). Specifically, our model explicitly considers within-ejaculate competition or competition among sperm from many males but not competition among sperm from pairs (or small numbers) of males. 
%n the introduction we mention that sperm competition has been shown to alter allele frequencies and fitness, and in this case we now mention this was shown within a single ejaculate (L68).
}

@@

Scott et al. present a 3-locus model for sex-determination (SD) evolution (1 locus has an ancestral SD function; 1 has a derived SD function; 1 locus affects selection in the haploid or diploid phase of the life cycle). Their results converge on previously-identified conditions for the spread of new SD alleles, while also broadening the range of scenarios and conditions that can favour transitions between SD systems.

The paper is interesting and conceptually appropriate for PLoS Biology. The authors have done a great job of analysing a somewhat complicated model and drawing out a relatively complete picture of the conditions favouring the evolution of new sex determination systems, and I expect their expert treatment will go over well with people that are close to the research context of the new theory.

On the other hand, I am much less sure that this paper will reach beyond experts in this subject area, though the authors should seek to broaden the appeal. This is not because the results are narrow -- I think they can have a broad appeal. Rather, the paper is written in a technical manner that is likely to limit the appeal beyond the expert class. I am reasonably comfortable with models like this, but still found the presentation to be very hard going in a variety of places, particularly in the results. I offer some suggestions for presentation that may help improve the presentation and broaden the potential appeal of the paper.

\textcolor{blue}{
We thank the reviewer for their thoughtful, positive, and very constructive comments. We have substantially restructured the Introduction, Model, and Results in response to these suggestions (and comments from other reviewers), as outlined below. 
This has significantly improved the accessibility and appeal of our results. 
}

@@

\noindent\subsection{}
I think the introduction could be simplified so that the context of the study is more obvious. The authors essentially need to make three key points, which are there, but could be better drawn out. First, transitions between sex-determining systems are common, and recent empirical work has given us a much greater appreciation for just how common these transitions are. Second, there are two major mechanisms that are thought induce evolution of new sex-determination mechanisms: (1) selection to stabilize the sex ratio, and (2) hitchhiking between (tightly linked) sex-determination and sexually antagonistic alleles. Third, the authors should clearly outline the major limitations of current theory as a means of motivating their mega-model that captures an array of genetic and selection scenarios that may lead to transitions in sex determination. As I mentioned, these elements can all be found in the introduction, but the pitch can be substantially streamlined to allow readers to easily see: (1) what is the context? (2) What do we currently think drives transitions between SD systems? And (3) what are the holes in the theory and how will the authors address them?

\noindent\paragraph{Response}
\textcolor{blue}{
We have simplified the introduction according to the suggested format. In addition, when discussing holes in current theory we have expanded our discussion of previous studies that include haploid selection, as suggested by reviewer 3. 
To accommodate these changes, we have moved the discussion of empirical evidence for haploid selection to the discussion (L441-454). The introduction is now structured as follows:
\begin{enumerate}
\item[1.]{The first paragraph discusses how common transitions are, and now mentions that recent empirical evidence is making this increasingly apparent. (L2-20)}
\item[2-3.]{The second and third paragraphs lay out the two major mechanisms thought to drive transitions in sex-determination, (1) selection to stabilize the sex ratio, and (2) linkage associations between alleles favoured in one sex and the alleles that determine that sex. (L21-29, L30-38) }
\item[4.]{Connecting to point (2), we then found it useful to introduce one novel aspect of our model -- we consider a selected locus that is very tightly linked to the ancestral sex-determining locus and find the surprising result that transitions in which linkage is reduced can be favoured. (L39-48)}
\item[5.]{Next, we introduce the most significant novel model feature, haploid selection, which we present as an evolutionary problem in light of (1) and (2). (L49-56) }
\item[6.]{The sixth paragraph then describes previous studies that have included haploid selection (see also reviewer 3 recommendations). These studies suggest that (1) and (2) are both very important with haploid selection and point to the major limitations in solving the evolutionary puzzle. (L57-69)}
\item[7.]{Finally, we lay out our main results in this context. This makes it clear that our general model is necessary to determine the important forces driving transitions in sex determination (e.g., considering various genomic locations). We particularly highlight the unexpected conclusions that our model allows us to reach. (L70-83)}
\end{enumerate}
}

@@

\noindent\subsection{}
Likewise, the results yield three major insights:\\
(1) SD transitions can, counterintuitively, potentially INCREASE sex ratio bias\\
(2) Tight linkage and sexual antagonism are not required for transitions to occur\\
(3) Under haploid selection, conditions for the spread of new sex-determining alleles are extremely permissive.\\
Again, all of these points are there, but they could be better drawn out than they currently are. The results, as written, are hard-going, and I expect will be particularly challenging for readers that are not heavily immersed in evolutionary theory.

One potential way to help readers -- particularly those interested in what the paper has to say, but are not mathematically inclined -- is to conspicuously highlight each important result and briefly place it into context of earlier theory. But at the same time, the authors have done a great job of drawing out general mathematical results from a relatively complicated model -- their mathematical insights ARE the results and I am not suggesting that they be minimized or shunted into appendices.

One tactic for including analytical details which specialists will want to see, while allowing less mathematically inclined readers to easily find each main result, is to place short and accessible summaries at the close of each section of the Results. Such a tactic has been put to good use in a hugely influential paper by Allen Orr (1998 Evolution), which I expect the authors will be familiar with. Orr presents the major mathematical results of his analysis alongside prominently labelled punchline paragraphs that summarize each of his main results in a concise and easily understandable manner. If the authors choose to take this approach, please be clear to place the take-home messages in context: clearly state whether the result confirms previous theory, or represents a new result or contradiction of an earlier result.

\noindent\paragraph{Response}
\textcolor{blue}{
This is an excellent suggestion and has greatly improved the clarity of presentation in our manuscript. In the results, we have specifically highlighted our main conclusions in the style of Orr 1998 (slightly modified from the three conclusions suggested by the reviewer): 
%We now give 3 main conclusions at the end of the introduction (L69) which echo those mentioned by the reviewer (looser linkage can evolve, equal selection to increase or decrease the sex ratio, ESD does not always invade biased sex ratio; all of which suggest haploid selection makes conditions for transitions more permissive).
\begin{enumerate}
\item[(1)] Selection on loci in or near the non-recombining region around the ancestral sex-determining locus ($r\approx0$) prevents cis-GSD transitions (XY $\leftrightarrow$ XY, ZW $\leftrightarrow$ ZW) but can spur trans-GSD transitions (XY $\leftrightarrow$ ZW). (L228-230)
\item[(2)] With tight linkage between a selected locus and the ancestral sex-determining locus ($r\approx0$), trans-GSD transitions (XY $\leftrightarrow$ ZW) can be favoured by selection \textit{even if they weaken sex-linkage} ($r<R$), potentially shifting sex determination to a different chromosome ($R=1/2$). 
Such transitions can also lead to the maintenance of multifactorial sex-determination systems. (L282-286)
\item[(3A)] New sex-determining alleles (causing cis- \textit{or} trans-GSD transitions) can spread if they arise more closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($R<r$). (L324-326)
\item[(3B)] New sex-determining alleles (causing trans-GSD transitions, XY $\leftrightarrow$ ZW) can spread even if they arise less closely linked with a locus that experiences selection than the ancestral-sex-determining locus is ($r<R$), requiring haploid selection. (L327-330)
\item[(4)] When selection is weak relative to recombination, the presence of haploid selection equally favors the spread of new sex determination systems that reduce sex-ratio bias (benefiting from Fisherian sex ratio selection) or generate a sex-ratio bias (benefiting from associations with selected alleles). (L350-353)
\item[(5)] Transitions from genetic to environmental sex-determination are not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present.  (L411-413)
%\vspace{0.25cm}\\
\end{enumerate}
These conclusions are also more clearly laid out in the introduction (L48-49, L74-85) and are referenced directly in summarisations in the discussion (L424-432, L543-554). 
%As (1A), (1B), and (2) have some antecedents in the literature, we discuss these immediately below each conclusion. 
%We also discuss, immediately below each conclusion, its novelty.
%These are all new results, although (1A) was shown for the special case of $r=0$, $R=1/2$, and ploidally-antagonistic selection in Kozielska et al 2010, which we now mention immediately below (1A).
\textcolor{red}{what is new and what is confirmatory? [see red notes in main text in conclusions]} 
}

\textcolor{blue}{
Furthermore, to reduce the technicality of our presentation, we have restructured the presentation of the results subsections "Generic invasion by a neo-Y or neo-W" and ``Tight linkage with the ancestral sex-determining locus''. 
Some details have been moved to the supplementary material (particularly, proofs regarding the leading eigenvalue, formally LXXX-XXX). 
In the ``Tight linkage'' section, we now first describe specific cases why neo-W's can have improved female fitness, L235-259, with more focus on non-technical interpretation (e.g., L237-240). 
Then we discuss the fact that looser sex-linkage can evolve (L260-275, Conclusion 2). 
}

@@

\noindent\subsubsection{}
Immediately preceding the paragraph beginning line 44: It would be worth establishing the limitations of previous work. What exactly motivated you to do this mathematical modelling?

\noindent\paragraph{Response}
\textcolor{blue}{We have rearranged and rewritten the introduction to better motivate our model (see Response 2.1). In particular, paragraph 5 presents the role of haploid selection as an evolutionary puzzle and paragraph 6 explains the limitations of previous attempts to solve the puzzle. (L49-69)}

\noindent\subsubsection{}
L56-57: Rephrase for clarity "individual cases of meiotic drive are generally sex-limited, and exclusively effect male or female gametes". And on line 57, "sex specific" could be changed to "sex-limited".

\noindent\paragraph{Response}
\textcolor{blue}{
Thank you for this suggestion, ``sex-limited'' is a more efficient way to communicate that haploid selection often only occurs in gametes produced by one sex. 
We now use ``sex-limited'' throughout, with two exceptions. L72 and L126, where we are referring to the fact that we model haploid selection in either/both sexes (focussing on sex-limited cases in our examples). 
This section was re-written but the section that roughly corresponds to the highlighted sentence now reads: ``...haploid selection is typically sex-limited in that it usually occurs among gametes produced by one sex only." (L51-52) 
}

@@

\noindent\subsubsection{}
Lines 80-81: This primary finding is completely aligned with the earlier views of theory that are summarized above. Are you sure you want to lead with this as the main emphasis of the study? It understates what is new.

\noindent\paragraph{Response}
\textcolor{blue}{
As suggested above, we have re-written to focus on our most novel findings, which are now laid our more specifically, e.g., L71-85 (see also Responses 2.1 and 2.2). 
}

@@

\noindent\subsubsection{}
M locus is dominant to X. How important is this assumption? Is this compatible with any data (e.g., on feminizing or masculinizing factors in Musca domestica)?

\noindent\paragraph{Response}
\textcolor{blue}{

This is an important assumption.
As discussed in the supplementary of van Doorn \& Kirkpatrick (2007, Nature), partial dominance and incomplete penetrance (partial epistatic dominance) reduce the range parameter values that favour transitions.
However, this reduction in parameter space is smooth and does not preclude transitions, meaning the qualitative conclusions are robust.
\textcolor{red}{I actually think that parameter space is only reduced because they also have a locus near the ancestral SDR that selects against new sex determining systems. In a 3-locus system, like ours, I would expect the $\lambda-1$ to be lower for incomplete penetrance (partial epistatic dominance) but not to change sign. }
Further, there are many systems in which one sex-determining allele is epistatically dominant.
And finally, our Mathematica file allows for arbitrary epistatic dominance coefficients.
We now discuss the empirical evidence for epistatic dominance, our general Mathematica file, and the analysis of van Doorn and Kirkpartick when introducing our model. (L94-105)
}
%\textcolor{red}{This is obviously very important, and is commonly assumed (Kozielska et al. 2010, van Doorn and Kirkpatrick 2007, 2010). There are a large number of other possibilities (given 9 genotypes that determine 2 sexes, ie $2^9$), and it seems like these alternatives could either increase or decrease selection on the invader. van Doorn and Kirkpatrick don't give any motivation (from my quick skim), but Kozielska et al. cite the housefly (Musca domestica), where it seems a single W is epistatically dominant over any number of loci with Y alleles. Is this ``compatible" with our model?}
%
%\textcolor{red}{Our analysis shows that the end state of the system depends on the `epistatic dominance' (term removed from our paper?) of the loci. Could draw this out more for this response. The drift-transitions Genetics paper rests on the difference in dominance, I'm pretty sure they have examples and the majority of new GSD systems are epistatically dominant. Also vD\&K 2010 say that there is evidence that epistatic dominance is common, and explain why a completely recessive should not invade. }

### 2018-05-03 LaTeX/main_revised.tex
more minor adjustments to Tight Linkage sections, beginning on response letter detail with new changes

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-specific in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-differences in diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-specific meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx} considered ancestral-ESD (with no sex-ratio bias) and numerically showed that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased.
This suggests that the benefits of associating with driving alleles overwhelms selection to balance the sex ratio and that haploid selection should cause increased sex-linkage. %, at least when new sex-determining alleles are rare. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} considered an ancestral GSD system that is perfectly linked to a meiotic driver (creating a sex ratio bias). 
They found that new, completely unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
This suggests that Fisherian sex-ratio selection overwhelms the benefits of being associated with driving alleles and that haploid selection should cause decreased sex-linkage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Model Outline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
%\centering
\centerline{\includegraphics[width=\linewidth]{../Plots/Sex_determination_outline.pdf}}
\caption{
{\bf Outline of model features.}
Panel A: Recombination rate parameters between the ancestral-sex-determining locus ($\mathbf{X}$, here assumed to have alleles X and Y), a locus under selection ($\mathbf{A}$, with alleles $A$ and $a$), and a new sex-determining locus ($\mathbf{M}$, with alleles $M$ and $m$). 
If $r<1/2$, then associations between ancestral sex-determining alleles and selected alleles can be present in new male gametes/gametophytes. 
Panel B: Haploid selection is often sex-specific, occurring during haploid production or competition in either males or females. 
For example, haploid selection in males only is represented by the dashed circle.  
If X or Y alleles remain associated with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X- or Y-bearing male gametes/gametophytes will be more abundant after haploid selection. 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determining systems. 
Similarly, zygotic sex-ratio biases can arise if haploid selected alleles are associated with new sex-determining alleles ($R<1/2$). 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW), a neo-Y allele spreads to pseudo-fixation (its maximum frequency among male gametes) and the ancestral Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele spreads to pseudo-fixation (its maximum frequency among female gametes) and the ancestral X allele is lost. 
Neo-W alleles allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%Essentially, neo-W alleles are able to invade (even with $R>r$) because the ancestral X experiences selection in males. 
%This can lead to (1) XX females having lower than maximum fitness ($\Lambda_{W'A}^{(XY)}>1$) and/or (2) the Y carrying alleles that are beneficial in females. 
\noindent Broadly, it is possible for neo-W females to have higher fitness than XX females for three reasons.
Firstly, because the ancestral X experiences selection in both males and females, it may be unable to specialize strongly on an allele favoured in females.
Secondly, an allele can be associated with the Y and yet favoured in females.
Thirdly, haploid selection can create a male sex-ratio bias. 
In turn, a neo-W can spread because (a) it is only found in females and is therefore unleashed from counterselection in males (corresponding to $\Lambda_{W'A}^{(XY)}>1$), (b) it allows Y-associated alleles into females (corresponding to $\Lambda_{W'a}^{(XY)}>1$), and (c) it may produce the rarer sex and thus be favoured by sex-ratio selection (increasing both haplotypic growth rates).
To clarify conditions under which trans-GSD transitions can occur, we focus here on cases where there is no haploid selection (and hence no sex-ratio bias) and discuss the additional effect of haploid selection in \nameref{app:inv_cond}.

@@

In some cases, both neo-W-$A$ and neo-W-$a$ haplotypes can spread.
For example, when $AA$ individuals have low fitness in females yet the $A$ is polymorphic or fixed on the X background due to overdominance in males (Fig \ref{fig:regionplots}B and \ref{fig:regionplots}C), both neo-W-$A$ and neo-W-$a$ haplotypes produce fewer unfit $AA$ females.
This is true for the neo-W-$A$ haplotype because it can pair with a Y-$a$ haplotype and still be female. 
Whenever both haplotypic growth rates are greater than one, invasion by a neo-W is expected regardless of its linkage with the selected locus (i.e., for any $R$, see \nameref{fig:positionOverdominance} and \nameref{fig:temporalOverdominance} for examples). 
As a consequence, evolution can favor a new sex determination system on a different chromosome ($R = 1/2$), despite the fact that this unlinks the sex-determining locus from the selected locus.

@@

\noindent
With haploid selection, Conclusions 1 \& 2 continue to apply (\nameref{app:inv_cond}). 
The parameters for which neo-W-$A$ and neo-W-$a$ haplotypes spread under various forms of haploid selection are plotted in \nameref{fig:regionMaleDrive}, \nameref{fig:regionMaleGS}, \nameref{fig:regionFemaleDrive}, \nameref{fig:regionFemaleGS}.
In particular, we note that adding haploid selection allows shifts in sex determination to a different chromosome ($R=1/2$) even when selection is sexually antagonistic with directional selection in each diploid sex, e.g., \nameref{fig:SexAntagTighterMaleDrive}.  
Furthermore, haploid selection allows variation to be maintained by ploidally-antagonistic selection, under which trans-GSD transitions may also be favoured, \nameref{fig:regionPloidAntag}.
Some cases of XY $\rightarrow$ ZW transitions where $r=0$, $R=1/2$, and selection is ploidally-antagonistic (meiotic drive in males opposed by diploid selection) were studied by Kozielska et al. \cite{Kozielska:2010vm}, who found that sex-ratio biases are reduced during these transitions. 
However, such transitions are not always driven by selection to reduce sex-ratio bias. 
For example, with XY sex determination and haploid selection in females, sex ratios are not ancestrally biased yet a neo-W can invade (\nameref{fig:regionPloidAntag}).

@@

When selection is weak relative to recombination, the absence of haploid selection ($t^\circ=\alpha^\circ_{\Delta}=0$), implies that the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W'}^{(XY)}=\lambda_{Y'}^{(XY)}$), such that trans-GSD transitions can also occur only if the new sex-determining locus is more closely linked to a locus under selection ($R<r$), as found by \cite{vanDoorn:2010hu}. 
When there is haploid selection, invasion also typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Fig \ref{fig:Combination_Centimorgans}).
For example, if the $\mathbf{A}$ locus is unlinked to the ancestral sex-determining locus ($r=1/2$, which implies that $ \hat{p}^\male_Y-\hat{p}^\male_X =0$), a more closely linked neo-W ($R<1/2$) can always invade. 
%In this case, there is no ancestral association between $\mathbf{A}$ alleles and sex chromosomes in males, $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$, see Eq \eqref{eq:freq_diffs}. 
%The second term in Eq \eqref{eq:lambda_neoW} therefore disappears and invasion depends only on the sign of $(r-R)$.%, as in the case of the neo-Y. 
However, with haploid selection and some ancestral sex-linkage ($r<1/2$; allowing allele frequency differences on the X and Y), the term in square brackets in Eq \eqref{eq:lambda_neoW} can be positive.
This implies the following two conclusions:

>>>>

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-limited in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-differences in diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-limited meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx} considered ancestral-ESD (with no sex-ratio bias) and numerically showed that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased.
This suggests that the benefits of associating with driving alleles overwhelms selection to balance the sex ratio and that haploid selection should cause increased sex-linkage. %, at least when new sex-determining alleles are rare. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} considered an ancestral GSD system that is perfectly linked to a meiotic driver (creating a sex ratio bias). 
They found that new, completely unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
This suggests that Fisherian sex-ratio selection overwhelms the benefits of being associated with driving alleles and that haploid selection should cause decreased sex-linkage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Model Outline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
%\centering
\centerline{\includegraphics[width=\linewidth]{../Plots/Sex_determination_outline.pdf}}
\caption{
{\bf Outline of model features.}
Panel A: Recombination rate parameters between the ancestral-sex-determining locus ($\mathbf{X}$, here assumed to have alleles X and Y), a locus under selection ($\mathbf{A}$, with alleles $A$ and $a$), and a new sex-determining locus ($\mathbf{M}$, with alleles $M$ and $m$). 
If $r<1/2$, then associations between ancestral sex-determining alleles and selected alleles can be present in new male gametes/gametophytes. 
Panel B: Haploid selection is often sex-limited, occurring during haploid production or competition in either males or females. 
For example, haploid selection in males only is represented by the dashed circle.  
If X or Y alleles remain associated with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X- or Y-bearing male gametes/gametophytes will be more abundant after haploid selection. 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determining systems. 
Similarly, zygotic sex-ratio biases can arise if haploid selected alleles are associated with new sex-determining alleles ($R<1/2$). 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW), a neo-Y allele spreads to pseudo-fixation (its maximum frequency among male gametes) and the ancestral Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele spreads to pseudo-fixation (its maximum frequency among female gametes) and the ancestral X allele is lost. 
Neo-W alleles allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%Essentially, neo-W alleles are able to invade (even with $R>r$) because the ancestral X experiences selection in males. 
%This can lead to (1) XX females having lower than maximum fitness ($\Lambda_{W'A}^{(XY)}>1$) and/or (2) the Y carrying alleles that are beneficial in females. 
\noindent 
To clarify conditions under which trans-GSD transitions can occur, we focus here on cases where there is no haploid selection (and hence no sex-ratio bias) and discuss the additional effect of haploid selection in \nameref{app:inv_cond}. 
Broadly, it is possible for neo-W females to have higher fitness than XX females for two reasons.
Firstly, because the ancestral X experiences selection in both males and females, it may be unable to specialize strongly on an allele favoured in females.
Secondly, an allele can be associated with the Y and yet favoured in females.
In turn, a neo-W can spread because (a) it is only found in females and is therefore unleashed from counterselection in males (corresponding to $\Lambda_{W'A}^{(XY)}>1$), (b) it allows Y-associated alleles into females (corresponding to $\Lambda_{W'a}^{(XY)}>1$).

@@

In some cases, both neo-W-$A$ and neo-W-$a$ haplotypes can spread.
For example, when $AA$ individuals have low fitness in females yet the $A$ is polymorphic or fixed on the X background due to overdominance in males (Fig \ref{fig:regionplots}B and \ref{fig:regionplots}C), both neo-W-$A$ and neo-W-$a$ haplotypes produce fewer unfit $AA$ females.
This is true for the neo-W-$A$ haplotype because it can pair with a Y-$a$ haplotype and still be female. 
Whenever both haplotypic growth rates are greater than one, invasion by a neo-W is expected regardless of its linkage with the selected locus (i.e., for any $R$), see \nameref{fig:positionOverdominance} and \nameref{fig:temporalOverdominance} for examples. 
As a consequence, evolution can favor a new sex determination system on a different chromosome ($R = 1/2$), despite the fact that this unlinks the sex-determining locus from the selected locus.

@@

\noindent
With haploid selection, Conclusions 1 \& 2 continue to apply (\nameref{app:inv_cond}). 
The parameters for which neo-W-$A$ and neo-W-$a$ haplotypes spread under various forms of haploid selection are plotted in \nameref{fig:regionMaleDrive}, \nameref{fig:regionMaleGS}, \nameref{fig:regionFemaleDrive}, \nameref{fig:regionFemaleGS}.
In particular, we note that adding haploid selection allows shifts in sex determination to a different chromosome ($R=1/2$) even when selection is sexually antagonistic with directional selection in each diploid sex, e.g., \nameref{fig:SexAntagTighterMaleDrive}.  
Furthermore, haploid selection allows variation to be maintained by ploidally-antagonistic selection, under which trans-GSD transitions may also be favoured, \nameref{fig:regionPloidAntag}.
Some cases of XY $\rightarrow$ ZW transitions where $r=0$, $R=1/2$, and selection is ploidally-antagonistic (meiotic drive in males opposed by diploid selection) were studied by Kozielska et al. \cite{Kozielska:2010vm}, who found that sex-ratio biases are reduced during these transitions. 
However, such transitions are not always driven by selection to reduce sex-ratio bias. 
For example, with XY sex determination and haploid selection in females, sex ratios are not ancestrally biased yet a neo-W can invade (\nameref{fig:regionPloidAntag}). 
We further discuss how the spread of neo-sex-determining alleles is influenced by associations with haploid selected loci in the next section.

@@

When selection is weak relative to recombination, the absence of haploid selection ($t^\circ=\alpha^\circ_{\Delta}=0$), implies that the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W'}^{(XY)}=\lambda_{Y'}^{(XY)}$), such that trans-GSD transitions can also occur only if the new sex-determining locus is more closely linked to a locus under selection ($R<r$), as found by \cite{vanDoorn:2010hu}. 
When there is haploid selection, invasion also typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Fig \ref{fig:Combination_Centimorgans}).
For example, if the $\mathbf{A}$ locus is unlinked to the ancestral sex-determining locus ($r=1/2$, which implies that $ \hat{p}^\male_Y-\hat{p}^\male_X =0$), a more closely linked neo-W ($R<1/2$) can always invade. Thus, 
%In this case, there is no ancestral association between $\mathbf{A}$ alleles and sex chromosomes in males, $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$, see Eq \eqref{eq:freq_diffs}. 
%The second term in Eq \eqref{eq:lambda_neoW} therefore disappears and invasion depends only on the sign of $(r-R)$.%, as in the case of the neo-Y.
